---
title: "Shiny example"
author: "Xijin Ge"
date: "2/10/2020"
output: 
    html_document:
      runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## An example of using Shiny in an R Markdown document.

We can embed interactive contents with control widgets in R Markdown documents by including Shiny Apps. Note this needs the runtime: shiny at the 7th line above.

## Prepare data
```{r}
library(ggplot2)
state.x77 <- data.frame(state.x77)   
sta <- cbind(state.abb, state.x77, state.region) 
colnames(sta)[1] <- "State"               
colnames(sta)[10] <- "Region"              
sta$Pop.Density <- sta$Population/sta$Area
```

## This is the first example provided

 Notice the $input$myCol that specifies the column, and the **aes_string**, instead of just aes, and "Region" instead of Region.

```{r, echo = FALSE}
library(ggridges)
library(ggplot2)
selectInput("myCol",
 "Choose a column", choices = colnames(sta)[2:11])

renderPlot({
  
ggplot(sta, aes_string(x = input$myCol, y = "Region", fill = "Region")) +
  geom_density_ridges() +
  theme_ridges() + 
  theme(legend.position = "none", 
        plot.title = element_text(hjust = 0.5),
        axis.title.x = element_text(hjust = 0.5), 
        axis.title.y = element_text(hjust = 0.5))
  
})
```

## This is the challenge code that I was having trouble with

One plot shows up, but we cannot control it!!!
```{r, echo = FALSE}
selectInput("myCol", "Choose a column", choices = colnames(sta)[2:11])

renderPlot({
  ggplot(sta, aes_string( x = input$myCol) ) + 
    geom_histogram()
})
```


**It is actually being control by the botton in the first app!** Try to switch the first app and both changes! 
If you want both of them control by one widgets, that's fine. But if you want indepedent control. We need to rename the widget to **"myCol2"**. 

## Name the control widgets from myCol to myCol2 and it works
```{r, echo = FALSE}
selectInput("myCol2", "Choose a column", choices = colnames(sta)[2:11])

renderPlot({
  ggplot(sta, aes_string( x = input$myCol2) ) + 
    geom_histogram()
})
```


## Another example
We are re-cycling our boxplot code in the main workshop document. 
```{r, echo = FALSE}
selectInput("myCol3", "Choose a column", choices = colnames(sta)[2:11])

renderPlot({
  ggplot(sta, aes_string( x = "Region", y = input$myCol3, fill = "Region") ) +
    geom_boxplot() +
      geom_jitter(position=position_jitter(0.2))
  
})
```



# Told ya it is easy!
